---
title: "Solow Modell"
author: "Sebastian Kranz"
date: "Tuesday, December 30, 2014"
output: html_document
exercise_tag: "## Exercise"
---

## Exercise 1: Load Penn World Tables

Laden Sie zunächst die Penn World Tables im Stata Format aus dem Internet:

http://www.rug.nl/research/ggdc/data/pwt/v80/pwt80.zip

Entpacken Sie die ZIP Datei in ihrem Uebungsblatt Ordner

```{r}
library(foreign)
library(dplyr)
dat = read.dta("pwt80.dta")
descr = view.stata.var(dat)
dat = mutate(dat,
  gdp_cap = cgdpe / pop,
  y = cgdpe / pop,
  k = ck / pop,
  ci = csh_i
)

wgi = read.csv("wgi.csv")
colnames(wgi)[2] = "countrycode"
d = left_join(dat, select(wgi,-country),by = c("countrycode","year"))
library(countrycode)
d$region = countrycode(d$countrycode,"wb","region")
d$continent = countrycode(d$countrycode,"wb","continent")

dat = d

```

## Aufgabe 1 
Determine the relationship between savings rates and per capita incomes in 1960
and 2000. To do this, you will need to download for all countries the variable ci and y in both 1960 and 2000 in the PWT7.1.

The variable labeled ci is the investment share of GDP (i.e. savings rate) and the variable labeled y is per capita GDP of a country expressed relative to the US level in percentage terms. Thus if y=50 in the data, it means that country i has a per capita GDP in 1960 that is ½ the US level in that year. Hand in a scatter plot for 1960 and 2000. 

```{r}
library(dplyr)
library(sktools)

d = filter(dat, year==1960 | year == 2000)
d = select(d,ci,y,year,country) 
library(rCharts)
rPlot(y ~ ci | year, data = d, type = 'point', color = 'country')



```

Also calculate and hand in the correlation coefficient for saving rates and income levels in 1960 and 2000.
```{r}
summarise(group_by(dat,year), cor.y.ci = cor(y,ci,use = "complete.obs"))
lm(y ~ ci, data=dat)
```


Generate a motion plot
```{r}
unique(dat$country)
countries = c("Russia","Germany","France","Greece","Spain","China","India","Poland","United States")
d = filter(dat, country %in% countries)
d = filter(dat, year >= 1996)
p = gvisMotionChart(data = d,idvar = "country",timevar = "year",xvar = "ci",yvar="gdp_cap", colorvar="region", sizevar="pop")
plot(p)


```

```{r}

p = gvisMotionChart(data = dat,idvar = "country",timevar = "year",xvar = "ci",yvar="gdp_cap", colorvar="continent", sizevar="pop")
plot(p)

```

```{r}
cat(paste0(colnames(d), collapse=" + "))
library(regtools)

d$YEAR = as.factor(d$year) 
reg1 =lm(gdp_cap ~  ci + ControlofCorruption + GovernmentEffectiveness + PoliticalStabilityNoViolence + RegulatoryQuality + RuleofLaw+year,data=d)

effectplot(reg1,numeric.effect = "25-75")

reg1 =lm(gdp_cap ~  ci + ControlofCorruption + GovernmentEffectiveness + PoliticalStabilityNoViolence + RegulatoryQuality + RuleofLaw+year+hc,data=d)


reg1 =lm(gdp_cap ~  ci + ControlofCorruption + PoliticalStabilityNoViolence + RegulatoryQuality + RuleofLaw+year+hc,data=d)

d = mutate(group_by(d,country), avg.savings = mean(ci, na.rm=TRUE))
reg1 =lm(gdp_cap ~  avg.savings + year+hc,data=d)

effectplot(reg1,numeric.effect = "10-90", show.ci=TRUE)

```

Let us analyse growth

```{r}
library(sktools)
library(tidyr)
library(regtools)

dl = left_join(expand(dat, countrycode,year),dat)
dat = arrange(dl, countrycode, year)

dol = do(group_by(dat, country),
        transform.for.growth.analysis(.,ahead=5, fixed.vars=c("year","country","countrycode","continent","region")))

dn = do(group_by(dat, country),
        transform.for.growth.analysis(.,ahead=5, overlapping=FALSE, fixed.vars=c("year","country","countrycode","continent","region")))

dg = dn

d = filter(dg, year>=1996)
cat(paste0(colnames(dg), collapse=" + "))

reg = lm(data = dg, gdp_cap_growth ~ pop_growth + gdp_cap_start + hc_start + csh_i_start+
         ControlofCorruption_start + PoliticalStabilityNoViolence_start + RegulatoryQuality_start + RuleofLaw_start+GovernmentEffectiveness_start)


effectplot(reg, show.ci=TRUE)
           
reg = lm(data = dg, gdp_cap_growth ~ pop_growth + csh_i_start+ gdp_cap_start + hc_start + year + region)

rd = regression.data(formula(reg), data=get.regression.data(reg))
effectplot(reg,rd, show.ci=TRUE)


reg = lm(data = dg, gdp_cap_growth ~ pop_growth + csh_i_start+ gdp_cap_start + hc_start + year + continent)

ereg = expanded.regression(reg)
effectplot(ereg, show.ci=TRUE)


```


Question 2: Calibrated Solow Model. Next calculate the average annual population growth rate
for each country between 1960 and 2000. If either the 1960 or 2000 data is missing, use the first
the years closest to these years for which data is available. For this you will need to download
the variable labeled POP in the PWT7.1. Now use the savings rate for 2000 for each country you
imported in Question 1. Assume all countries have the same depreciation rate δ= .05, the same
rate of exogenous technological change, γ=.02, and the same TFP parameter A=1. Use the
formula for the ratio of the balanced growth path per capita output for the Solow model that we
used in the 2/10/2014 class BGP
US t
BGP
i t y y , , / , where the comparison is being made to the US. More
specifically, take the average savings rate and average annual population growth rate for each
country and compute its output per capita relative to the US level.
Hand in a scatter plot which relates Solow relative GDPs per capita BGP
US t
BGP
i t y y , , 100 / and actual
relative 2000 GDPs per capita (i.e. y in the data). We multiply Solow relative GDP pre capita by
100 so as to put it in percentage terms as is the case with the variable y.
Note: On Friday, Seyed teach you a few tricks for downloading and importing the data into
excel.